---
title: 'Reproducible Research: Peer Assessment 1'
output:
  word_document: default
  pdf_document: default
  html_document:
    keep_md: yes
---

<!-- Set the working directory first. Make sure it is in the same folder of the .Rmd script-->
## Loading and preprocessing the data
```{r loadData, echo = TRUE}
#loading data
data<-read.csv("activity.csv")
#converting date column from factor to date type
data$date<-as.Date(data$date)
``` 


## What is mean total number of steps taken per day?
```{r createDataFrame, echo=TRUE}
#getting steps mean and median by date
steps_per_day_mean<-aggregate(data$steps, list(data$date), FUN = mean)
steps_per_day_median<-aggregate(data$steps, list(data$date), FUN = median)
steps_per_day_total<-aggregate(data$steps, list(data$date), FUN=sum)
#naming columns
names(steps_per_day_mean)<-c("date","stepsMean")
names(steps_per_day_median)<-c("date","stepsMedian")
names(steps_per_day_total)<-c("date","stepsTotal")
#merging dataframes
steps_per_day<-merge(steps_per_day_mean,steps_per_day_median)
steps_per_day<-merge(merge(steps_per_day_mean,steps_per_day_median),steps_per_day_total)
```
####**Histogram of total steps per day**
```{r histPlot, fig.height=4}
library(ggplot2)
hist<-ggplot(steps_per_day,aes(x=steps_per_day$stepsTotal))
hist+geom_histogram(breaks=seq(0,25000,5000), aes(fill=..count..))+xlab("Total Steps")+ggtitle("Total Steps per day")
```
```{r echo=FALSE, results='hide'}
print("\n") 
```  
####**Mean and median total number of steps taken per day**
```{r mean_median, results='markup'}
steps_per_day[,c(1:3)]
```

## What is the average daily activity pattern?
####**Time series plot of the 5-minute interval and the average number of steps taken, averaged across all days**
```{r}
steps_per_int_mean=aggregate(data$steps, list(data$interval), FUN=mean, na.rm=TRUE)
names(steps_per_int_mean)<-c("interval","stepsMean")
```
```{r, fig.height=4}
plot(steps_per_int_mean, type="l",col="blue")
title("Steps Average per interval")
```
```{r echo=FALSE, results='hide'}
print("\n") 
```  

####**5-minute interval containing the maximum number of steps**

```{r}
steps_per_int_mean[which(steps_per_int_mean$stepsMean==max(steps_per_int_mean$stepsMean)),1]
```
##**Imputing missing values**
####**Number of missing values in the dataset**
```{r}
nrow(data[ is.na(data$steps),])
```
####**Filling the missing values**
Missing values were filled using the median of the steps on all other days
```{r}
dataFilled<- data
steps_per_int_median=aggregate(data$steps, list(data$interval), FUN=median, na.rm=TRUE)
names(steps_per_int_median)<-c("interval","stepsMedian")

#Filling na values in steps field with steps median per interval

for (i in seq(1,nrow(dataFilled))){
        int <- dataFilled$interval[i]
        if  (is.na(dataFilled$steps[i])){
                dataFilled[1][i,] <- steps_per_int_median[steps_per_int_median$interval == int,]["stepsMedian"]
        }
}

```
They were then calculated the total, average and median steps using the new dataset
```{r}
steps_per_day_mean_DF<-aggregate(dataFilled$steps, list(dataFilled$date), FUN=mean)
steps_per_day_median_DF<-aggregate(dataFilled$steps, list(dataFilled$date), FUN=median)
steps_per_day_total_DF<-aggregate(dataFilled$steps, list(dataFilled$date), FUN=sum)
names(steps_per_day_mean_DF)<-c("date","stepsMean")
names(steps_per_day_median_DF)<-c("date","stepsMedian")
names(steps_per_day_total_DF)<-c("date","stepsTotal")
steps_per_day_DF<-merge(merge(steps_per_day_mean_DF,steps_per_day_median_DF),steps_per_day_total_DF)
```
The following histogram was made to present the results:
```{r histPlot1, fig.height=4}
library(ggplot2)
hist<-ggplot(steps_per_day_DF,aes(x=steps_per_day_DF$stepsTotal))
hist<-hist+geom_histogram(breaks=seq(0,25000,5000), aes(fill=..count..))+xlab("Total Steps")+ggtitle("Total Steps per day")
hist<-hist +scale_fill_gradient("Count", low = "#093B0D", high = "#98D69D")
hist

```
```{r echo=FALSE, results='hide'}
print("\n") 
```  
As you can see the values differ from the first histogram for the first five thousand steps. 


## Are there differences in activity patterns between weekdays and weekends?
####**Creating a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.**
```{r}
library(plyr)
Sys.setlocale("LC_TIME", "C") # setting LOCAL TIME -> English
dataFilled<-mutate(dataFilled, dayType = ifelse(weekdays(date) %in% c("Saturday","Sunday","sábado","domingo"),"weekend","weekday"))
```
####**Making a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)**
```{r fig.height=4}
DFSummary<-aggregate(dataFilled$steps, by=list(dataFilled$dayType,dataFilled$interval),FUN=mean)
names(DFSummary)<-c("DayType","Interval","Steps")
g<-ggplot(DFSummary,aes(Interval,Steps))
g+geom_line(color="#017292") +facet_wrap(~DayType, nrow=2, ncol=1)+labs(x="Interval")+labs(y="Number of steps")+labs(title="Steps by day type")
```
```{r echo=FALSE, results='hide'}
print("\n") 
```  
As you can see, there are some differences between the patterns of weekday and weekend day

